---
title: "Foundations of Probability and Statistics"
author: "Pasquale Gravante & Antonio Mastroianni"
date: "2022-10-24"
output: 
  html_document:
    df_print: paged
    code_folding: hide
    toc: true
    number_sections: true
---

The dataset consists of data regarding a sample consisting of  **299 patients** possibly affected by cardiovascular deseases.
The variables taken into account for the analysis are:

*   **Age**: integer representing the age of the patient
*   **Sex**: sex of the patient **(Boolean)**
*   **Anaemia**: decrease of red blood cells or hemoglobin **(Boolean)**
*   **Creatinine_phosphokinase**: level of the CPK enzyme in the blood **(mcg/L)**
*   **Diabetes**: whether the patient has diabetes **(Boolean)**
*   **Ejection_fraction**: Percentage of blood leaving the heart at each contraction
*   **high_blood_pressure**: whether the patient suffers from hypertension
*   **Platelets**: Platelets in the blood **(kiloplatelets/mL)*
*   **serum_creatinine**: Level of serum creatinine in the blood **(mg/dL)**
*   **serum_sodium**: Level of serum sodium in the blood **(mEq/L)**
*   **Smoking**: If the patient smokes or not **(Boolean)**
*   **Time**: Follow-up period **(days)**
*   **Death_event**: If the patient died for a cardiac arrest or not **(Boolean)**

```{r Libraries, echo=TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyverse)
library(tidyr)
library(gridExtra)
library(GGally)
library(moments)
library(corrplot)
```
First of all, we decided to transform our categorical data into factors since they were recognized as integers.

```{r}
heart <- read.csv("~/Downloads/heart_failure_clinical_records_dataset.csv")
glimpse(heart)
```

The table shows the dataframe after transformation.

```{r Editing column type}
for (i in c(2,4,6,13)){
  heart[,i] = as.factor(heart[,i])
}
heart$sex[heart$sex==0] <- 'Female'
heart$sex[heart$sex==1] <- 'Male'
heart$smoking[heart$smoking==0] <- 'No'
heart$smoking[heart$smoking==1] <- 'Yes'
heart$sex = as.factor(heart$sex)
heart$smoking = as.factor(heart$smoking)
glimpse(heart)
```

# EDA

The table represents all indices for each quantitative variable.

```{r Data summary}
data.frame(unclass(summary(heart[,-c(2,4,6,10,11,13)])), check.names = FALSE, stringsAsFactors = FALSE)
```

## Measures of central tendency

The age histogram consists of 6 intervals, each of which has length 10. Density is represented on the y axis. Mean is represented by a red line.

```{r Age histogram}
ggplot(heart, aes(x = age)) +
  geom_histogram(aes(y = ..density..), binwidth = 10, fill = 'cyan3', color = 'white') +
  theme_minimal() + 
  geom_vline(xintercept = mean(heart$age), color = 'deeppink2') +
  labs(x = 'Age', y='Density')+
  scale_x_continuous(breaks= c(35, 45,55,65,75,85,95))
```

A comparison is made for the variable "creatinine_phosphokinase", which represents the concentration level for the enzyme CPK in the blood for male and females.

```{r Central Tendency}
ggplot(heart, aes(x = creatinine_phosphokinase, color = sex))+
  stat_ecdf()+
  labs(x = 'CPK', y='F(CPK)')
sex <- c('Male','Female')
mean <- c(mean(filter(heart, sex == 'Male')$creatinine_phosphokinase),mean(filter(heart, sex == 'Female')$creatinine_phosphokinase))
median <- c(median(filter(heart, sex == 'Male')$creatinine_phosphokinase),median(filter(heart, sex == 'Female')$creatinine_phosphokinase)
)
data.frame(sex, mean, median)
```
CPK concentration is higher in male with respect to females. 
This is proven by the values for the means in the table above.

The following histogram shows the impact of outliers for the mean (red). The meadian(purple) instead is not affected by the presence of outliers. 

```{r}
ggplot(heart, aes(x = creatinine_phosphokinase)) +
  geom_histogram(aes(y = ..density..), binwidth = 100, fill = 'cyan3', color = 'white') +
  theme_minimal() + 
  geom_vline(xintercept = mean(heart$creatinine_phosphokinase), color = 'deeppink2') +
  geom_vline(xintercept = median(heart$creatinine_phosphokinase), color = 'purple') +
  labs(x = 'CPK', y='Density')
```

## Variability measures

The following table shows the values for variance, standard deviation, range of variation, interquartile range and median absolute deviation of the CPK concentration for males and females.

```{r Variability measures, message = FALSE, warning = FALSE}
attach(heart)
var <- c(var(filter(heart, sex == 'Male')$creatinine_phosphokinase),
var(filter(heart, sex == 'Female')$creatinine_phosphokinase))
sd <- c(sd(filter(heart, sex == 'Male')$creatinine_phosphokinase),
sd(filter(heart, sex == 'Female')$creatinine_phosphokinase))
RoV <- c(diff(range(filter(heart, sex == 'Male')$creatinine_phosphokinase)),
diff(range(filter(heart, sex == 'Female')$creatinine_phosphokinase)))
interquartile_range <- function(x){
diff(quantile(x, probs = c(0.25, 0.75)))
}
iq <- c(interquartile_range(filter(heart, sex == 'Male')$creatinine_phosphokinase),
interquartile_range(filter(heart, sex == 'Female')$creatinine_phosphokinase))
MAD <- function(x) {
median(abs(x - median(x)))
}
mad <- c(MAD(filter(heart, sex == 'Male')$creatinine_phosphokinase),
MAD(filter(heart, sex == 'Female')$creatinine_phosphokinase))
data.frame(sex, var, sd, RoV, iq, mad)
```

## Visualizations

The following boxplots show the variability of Serum-Creatinine, CPK, platelets concentration and ejetion fraction according to a categorical variable.

```{r Boxplots fumatori}
bp5 = ggplot(heart, aes(x = smoking, y = serum_creatinine, fill = smoking)) + geom_boxplot() 
bp6 = ggplot(heart, aes(x = smoking, y = creatinine_phosphokinase, fill = smoking)) + geom_boxplot() 
bp7 = ggplot(heart, aes(x = smoking, y = platelets, fill = smoking)) + geom_boxplot() 
bp8 = ggplot(heart, aes(x = smoking, y = ejection_fraction, fill = smoking)) + geom_boxplot() 
grid.arrange(bp5,bp6,bp7,bp8, nrow = 2)
```

```{r Boxplots diabetici}
bp9 = ggplot(heart, aes(x = diabetes, y = serum_creatinine, fill = diabetes)) + geom_boxplot() 
bp10 = ggplot(heart, aes(x = diabetes, y = creatinine_phosphokinase, fill = diabetes)) + geom_boxplot() 
bp11 = ggplot(heart, aes(x = diabetes, y = platelets, fill = diabetes)) + geom_boxplot() 
bp12 = ggplot(heart, aes(x = diabetes, y = ejection_fraction, fill = diabetes)) + geom_boxplot()
grid.arrange(bp9,bp10,bp11,bp12, nrow = 2)
```

From the boxplots it can be deduced that people who smoke and have diabetes generally have lower values for the ejection fraction. This is generally considered a possible cause for Cardiomyopathy.

```{r 3rd and 4th moment }
heart$cpk_standard <- scale(heart$creatinine_phosphokinase)
mean(heart$cpk_standard)
d1 <- ggplot(filter(heart, sex == 'Male'), aes(x= cpk_standard, color = sex ))+
  geom_density(color = '#6dca7c')+
  labs(x='CPK', title = 'CPK concentration for males')+
  theme_bw()+
    scale_x_continuous(breaks= c(-1,0,1,2,3,4,5,6,7))
d2 <- ggplot(filter(heart, sex == 'Female'), aes(x= cpk_standard ))+
  geom_density(color = '#933183')+
  labs(x='CPK', title = 'CPK concentration for females')+
  theme_bw()+
  scale_x_continuous(breaks= c(-1,0,1,2,3,4))
grid.arrange(d1, d2)
#add labels
skew <- c(skewness(filter(heart, sex == 'Male')$creatinine_phosphokinase), skewness(filter(heart, sex == 'Female')$creatinine_phosphokinase))
kurt <- c(kurtosis(filter(heart, sex == 'Male')$creatinine_phosphokinase), kurtosis(filter(heart, sex == 'Female')$creatinine_phosphokinase))
data.frame(sex, skew, kurt)
```

In order to understand which variables might be related a paired scatterplot is represented.

```{r Paired Scatterplot}
ggpairs(heart, columns = c(1,3,5,7,8,9), aes(color = sex), diag = list(continuous = 'blankDiag'), upper = list(continuous = 'points'))+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))
```

## Chi-squared

```{r Contingency table and chi-square}
classes = cut(heart$ejection_fraction, c(14, 30, 60, 80))
classes = as.character(classes)
heart$ejection_fraction_classes = classes
heart$ejection_fraction_classes[heart$ejection_fraction_classes == "(14,30]"] = "Low"
heart$ejection_fraction_classes[heart$ejection_fraction_classes == "(30,60]"] = "Medium"
heart$ejection_fraction_classes[heart$ejection_fraction_classes == "(60,80]"] = "High"
heart$ejection_fraction_classes = as.factor(heart$ejection_fraction_classes)
ft1 = ftable(heart$sex, heart$ejection_fraction_classes, heart$DEATH_EVENT)
ft1
chisq = chisq.test(ft1)
chisq.test(ft1)
chisq$residuals
```

```{r Qualitative Variables - Barplot}
  
ggplot(heart, aes(x = smoking, fill = sex))+
  geom_bar()+
  labs(x = 'Smoking', y='Frequency', title = 'Total of smoking males and females')
```

In order to understand which variables might be related a paired scatterplot is represented.

Apparently, there is not any relation between variables. 
Let's get a look at the numbers to check if this is true. 

```{r Variance covariance matrix}
as.data.frame(cov(heart[,c(1,3,5,7,8,9)]))
```

The correlationplot shows the absence of strong correlations between the variables taken into account. 

In order to understand which variables might be related a paired scatterplot is represented.

```{r Paired Scatterplots}
ggpairs(heart, columns = c(1,3,5,7,8,9), aes(color = sex), diag = list(continuous = 'blankDiag'), upper = list(continuous = 'points'))+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))
```

Scatters are colored according to "sex" variable. Apparently, there is not any relation between variables. 
Let's get a look at the numbers to check if this is true. 


```{r Correlation matrix}
corrplot(cor(heart[,c(1,3,5,7,8,9)]), method = 'color', addCoef.col = '#4d4c4b', order ='AOE', diag = FALSE, cl.cex = 0.5, tl.col='Black', tl.cex=0.8, tl.srt=70 ,)
```

The correlationplot shows the absence of strong correlations between the variables taken into account.
THe highest(negative) correlation value is between 'serum_creatinine' and 'serum_sodium'.

# Linear Regression

Let's give a look at the regression line for these two variables. 


```{r Linear Regression}
ggplot(heart, aes(x = serum_sodium, y = serum_creatinine, color = sex))+
  geom_point()+
  geom_smooth(method="lm", se = FALSE, color = 'deeppink3')
```


```{r Parameters}
reg <-lm(formula = heart$serum_creatinine ~ heart$serum_sodium)
as.data.frame(reg$coefficients)
```

## R-squared

While the goodness of fitness is equal to:
```{r Goodness of fitness}
cor(heart$serum_creatinine, heart$serum_sodium)^2
```
R^2 value is very low. Only 3.5% of the total variability of the data is explained by the displayed regression model.

# Inference

Let's take into account the CPK concentration variable. We want to know, based on the observations in the sample, what is the confidence interval containing the mean of the population.

We are going to perform a t-test because the variance of the population is not known.
We are going to consider a confidence level of 0.99

```{r Confidence Intervals}
t.test(heart$creatinine_phosphokinase, alternative = "two.sided", conf.level = 0.99)
```

The mean of the population resides in the interval [436.37; 727.30] with a confidence level of 0.99.

Let's assume to reduce our confidence level to 0.95.

```{r}
t.test(heart$creatinine_phosphokinase, alternative = "two.sided", conf.level = 0.95)

```

As expected, the confidence interval narrows as the confidence level decreases.


Let's also assume the case in which the analyzed sample has a smaller dimension.
In order to do that, we are going to sample without replacement our data and get 150 observation.

```{r Bootstrap Confidence interval}
h <- sample(heart$creatinine_phosphokinase, 150, replace = FALSE)
t.test(h, alternative = "two.sided", conf.level = 0.95)


```
As expected, when the sample dimension is reduced, the confidence interval increases.

# Anova

```{r}
attach(heart)
plot(creatinine_phosphokinase~ejection_fraction_classes)
plot.design(creatinine_phosphokinase~ejection_fraction_classes)
plot.design(creatinine_phosphokinase~ejection_fraction_classes, fun = median)
```

```{r}
par(mfrow=c(2,2))
qqnorm(creatinine_phosphokinase[ejection_fraction_classes=="Low"])
qqline(creatinine_phosphokinase[ejection_fraction_classes=="Low"])
qqnorm(creatinine_phosphokinase[ejection_fraction_classes=="Medium"])
qqline(creatinine_phosphokinase[ejection_fraction_classes=="Medium"])
qqnorm(creatinine_phosphokinase[ejection_fraction_classes=="High"])
qqline(creatinine_phosphokinase[ejection_fraction_classes=="High"])
```

```{r}
oneway.test(creatinine_phosphokinase~ejection_fraction_classes, var.equal = TRUE)
```
Since the p-value is greater than 0.05 we fail to reject the null hypothesis and so we can say that the means are equal.
